---
title: "Analyzing Music Data"
author: "fasehab"
format: html
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
```


## Introduction

This is a Quarto document for analyzing music data.

```{r}
billboard |> 
  select(artist, track, date.entered, wk1, wk2, wk3, wk4)
```

```{r}
n_values <- sum(!is.na(billboard$wk6))

ggplot(billboard, aes(x = wk6)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Week 6 Rankings",
    subtitle = paste("Number of values used:", n_values),
    x = "Week 6 Rank",
    y = "Count"
  )
```

## Missing vs. Present Rankings

```{r}
billboard |> 
  select(wk1, wk4, wk10, wk20, wk40, wk76) |> 
  pivot_longer(
    cols = everything(),
    names_to = "week",
    values_to = "rank"
  ) |> 
  mutate(week = factor(week, levels = c("wk1", "wk4", "wk10", "wk20", "wk40", "wk76"))) |> 
  group_by(week) |> 
  summarise(
    present = sum(!is.na(rank)),
    missing = sum(is.na(rank))
  )
```

## Re-entering Songs

We can check if any songs leave the chart and later re-enter it by comparing the actual number of weeks they were on the chart to the total span between their first and last appearance. If a song's total span of weeks is greater than the number of weeks it actually had a ranking, it must have left the chart and re-entered.

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank"
  ) |> 
  mutate(week = as.integer(str_remove(week, "wk"))) |> 
  filter(!is.na(rank)) |> 
  group_by(artist, track) |> 
  summarise(
    first_wk = min(week),
    last_wk = max(week),
    weeks_on_chart = n(),
    expected_weeks = last_wk - first_wk + 1,
    re_entered = weeks_on_chart < expected_weeks,
    .groups = "drop"
  ) |> 
  count(re_entered)
```

## Song Rankings Over Time

We reshape the `billboard` dataset to convert the week columns (`wk1` to `wk76`) into a long format with two columns: `week` and `rank`. Then, we plot each song's ranking over time.

```{r}
billboard_long <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(week = as.integer(str_remove(week, "wk")))

ggplot(billboard_long, aes(x = week, y = rank, group = interaction(artist, track))) +
  geom_line(alpha = 0.25) +
  scale_y_reverse() +
  labs(
    title = "Song Rankings Over Time",
    x = "Week",
    y = "Rank"
  )
```

## Song-Level Summary and Notable Songs

We construct a song-level summary displaying:
- **First rank**: The song's rank when it entered the chart.
- **Best rank**: The song's peak rank (minimum rank number).
- **First week at its best rank**: The week it first reached its peak rank.
- **Total weeks on the chart**: The total number of weeks the song remained on the chart.

We then identify three notable top-10 songs:
1. **The fastest song to reach #1** (minimum week to reach rank 1).
2. **The slowest song to reach #1** (maximum week to reach rank 1).
3. **The top-10 song with the longest chart run** (maximum total weeks on chart for a song whose best rank was <= 10).

```{r}
song_summary <- billboard_long |> 
  group_by(artist, track) |> 
  summarise(
    first_rank = rank[week == min(week)][1],
    best_rank = min(rank),
    first_week_at_best_rank = min(week[rank == best_rank]),
    total_weeks_on_chart = n(),
    .groups = "drop"
  )

# Display the first few rows of the summary
head(song_summary)

# 1. Fastest song to reach number one
song_summary |> 
  filter(best_rank == 1) |> 
  filter(first_week_at_best_rank == min(first_week_at_best_rank))

# 2. Slowest song to reach number one
song_summary |> 
  filter(best_rank == 1) |> 
  filter(first_week_at_best_rank == max(first_week_at_best_rank))

# 3. Top-10 song with the longest chart run
song_summary |> 
  filter(best_rank <= 10) |> 
  filter(total_weeks_on_chart == max(total_weeks_on_chart))
```

## Highlighting Notable Top-10 Songs

We filter the dataset to include only songs that reached the top 10 (peak rank <= 10). We display these songs in gray, while highlighting the three notable songs identified in the previous exercise with distinct colors and a legend.

```{r}
#| cache: true
top10_songs <- song_summary |> 
  filter(best_rank <= 10) |> 
  select(artist, track)

billboard_top10 <- billboard_long |> 
  semi_join(top10_songs, by = c("artist", "track")) |> 
  mutate(
    highlight = case_when(
      artist == "Madonna" & track == "Music" ~ "Madonna - Music (Fastest #1)",
      artist == "Lonestar" & track == "Amazed" ~ "Lonestar - Amazed (Slowest #1)",
      artist == "Creed" & track == "Higher" ~ "Creed - Higher (Longest Run)",
      TRUE ~ "Other Top-10 Songs"
    )
  )

background_data <- billboard_top10 |> filter(highlight == "Other Top-10 Songs")
highlight_data <- billboard_top10 |> filter(highlight != "Other Top-10 Songs")

ggplot() +
  geom_line(
    data = background_data,
    aes(x = week, y = rank, group = interaction(artist, track)),
    color = "gray80",
    alpha = 0.5,
    linewidth = 0.5
  ) +
  geom_line(
    data = highlight_data,
    aes(x = week, y = rank, color = highlight, group = interaction(artist, track)),
    linewidth = 1.2
  ) +
  scale_y_reverse() +
  scale_color_manual(
    values = c(
      "Madonna - Music (Fastest #1)" = "tomato",
      "Lonestar - Amazed (Slowest #1)" = "dodgerblue",
      "Creed - Higher (Longest Run)" = "forestgreen"
    ),
    name = "Notable Songs"
  ) +
  labs(
    title = "Rankings of Top-10 Songs Over Time",
    subtitle = "Highlighting three notable songs",
    x = "Week",
    y = "Rank"
  )
```

## Reading Music Data

We read and print the `music.csv` dataset using the `readr` package. According to the dataset documentation, `artist.familiarity` measures how familiar the artist is to listeners on a scale of 0 to 1, while `artist.hotttnesss` measures the artist's popularity on a scale of 0 to 1 at the time of data download (December 2010).

```{r}
#| message: false
music <- read_csv("data/music.csv")
music |> 
  select(artist.familiarity, artist.hotttnesss, song.year, song.tempo) |> 
  summary()
```

## Distribution of Song Years

We plot the distribution of the release years for the songs in the dataset. Note that some songs have their year recorded as `0` in the database; these are filtered out before plotting.

```{r}
music_clean <- music |> 
  filter(song.year != 0)

n_songs <- nrow(music_clean)

ggplot(music_clean, aes(x = song.year)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Song Years",
    subtitle = paste("Number of songs used:", n_songs),
    x = "Year",
    y = "Count"
  )
```

## Human-Readable Song Data

We select and print a subset of the dataset containing the most human-readable columns: artist name, location, coordinates (latitude and longitude), genre terms, familiarity, hotness, song title, and year.

```{r}
music |> 
  select(
    artist.name,
    artist.location,
    artist.latitude,
    artist.longitude,
    artist.terms,
    artist.familiarity,
    artist.hotttnesss,
    song.title,
    song.year
  ) |> 
  head()
```

## Counting Placeholder Values

Many columns in the dataset use placeholder values (typically `0` for numeric columns) to represent missing data. We count the number of placeholder values (`0`) in each of the selected columns:

```{r}
music |> 
  summarise(
    artist.location = sum(artist.location == 0),
    release.name = sum(release.name == 0),
    song.title = sum(song.title == 0),
    song.year = sum(song.year == 0),
    artist.familiarity = sum(artist.familiarity == 0),
    artist.hotttnesss = sum(artist.hotttnesss == 0)
  ) |> 
  pivot_longer(
    cols = everything(),
    names_to = "column",
    values_to = "placeholder_count"
  )
```

## Artist Locations Map

Only 3,742 out of 10,000 artists have usable location coordinates, while the remaining 6,258 have placeholder coordinates of (0, 0), which makes mapping them impossible.

```{r}
# Count usable vs placeholder coordinates
music |> 
  summarise(
    placeholder = sum(artist.latitude == 0 & artist.longitude == 0),
    usable = sum(artist.latitude != 0 | artist.longitude != 0)
  ) |> 
  pivot_longer(
    cols = everything(),
    names_to = "coordinate_status",
    values_to = "count"
  )

# Plot the map
library(maps)
world_map <- map_data("world")

ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "gray90", color = "white") +
  geom_point(
    data = filter(music, (artist.latitude != 0 | artist.longitude != 0) & song.year != 0),
    aes(x = artist.longitude, y = artist.latitude, color = song.year),
    alpha = 0.6,
    size = 0.8
  ) +
  scale_color_viridis_c(option = "plasma", name = "Year") +
  theme_minimal() +
  labs(
    title = "Geographic Distribution of Artist Locations by Song Year",
    subtitle = "Color indicates the release year for 1,825 artists with usable coordinates and non-zero years",
    x = "Longitude",
    y = "Latitude"
  )
```

